---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data



###load data
 The First real task is the unzip the data and load it into R
 
```{r  echo = TRUE  }
unzip("activity.zip")
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
```
 
## What is mean total number of steps taken per day?

### **Mean ** Total number of steps taken per day
  We will use the plyr and dplyr  package to work with our data. The plyr package must be loaded first.
  
```{r echo = TRUE}
library(plyr)
library(dplyr)
grp <- group_by(data,date) %>%
  summarize( total = sum(steps, na.rm=TRUE), nosteps = mean(steps,na.rm=TRUE) )
grp
```


### Make a histogram of the number of steps taken each day
  View a Histogram of the steps taken.
  
```{r echo = TRUE}
  
library(ggplot2)
qplot(grp$total, geom="histogram",main="Total Number of Steps Per Day")
  
```
  


### Calculate the mean and the median 
This is a calcuation on the number of steps taken per day

```{r echo = TRUE}
mean_num_of_steps <- mean(x = grp$total, na.rm = TRUE)
median_num_of_steps <- median(x= grp$total, na.rm = TRUE)
```

We find that the mean is `r mean_num_of_steps`     while the the median is found to be `r median_num_of_steps`





## What is the average daily activity pattern?

```{r echo = TRUE}
tm <- group_by(data,interval) %>%
  summarize(average = mean(steps,na.rm = TRUE), countsteps = sum(steps,na.rm=TRUE))

plot.ts(x = tm$interval, y=tm$average,type = "l", xlab = "5 min intervals", ylab = "Average across all days", main="Average daily activity pattern")

```
  
The 5-minute interval across all days that contains the maximum number of steps is: `r tm[which.max(tm$average),][[1]]`

The number of missings values in the dataset is: `r sum(is.na(data$steps))`
```{r echo=TRUE}
missing_NA_in_data <- sum(is.na(data$steps))
```



## Imputing missing values
We will use the mean of all steps accross all the weeks to apply to NA values in the dataset

```{r echo=TRUE}
# change interval type to factor
data$interval <- as.factor(data$interval)


fn_mean <- function(x){
  replace(x, is.na(x), mean(x,na.rm=TRUE))
}
newdata <- ddply(data, .(interval), transform,steps = fn_mean(steps))
 
head(newdata,30)



```


Here is a histogram of the dataset
```{r echo=TRUE}
newdata$interval <- as.character(newdata$interval)

newhistogram <- group_by(newdata,date) %>%
             summarise(total = sum(steps) )

#Historgram
qplot(newhistogram$total, geom="histogram",main="Total Number of Steps Per Day")


```

### central tendency

* Mean   : `r mean(newhistogram$total)`
* Median : `r median(newhistogram$total)`

The histogram differs from the first on on the zero x axis.  Those values are now located to the respective interval sections.

* Total With NAs:  `r sum(grp$total)`
* Total without NAs: `r sum(newhistogram$total)`



## Are there differences in activity patterns between weekdays and weekends?

For this section we will need to install a package *install.packages("chron")*  this is not place in the code area, as this would affect the performance when developing this document.  


```{r echo=TRUE}
#install.packages("chron")
library(chron)
newhistogram$date <- as.character.Date(newhistogram$date)


newdata$interval <- as.integer(newdata$interval);

wkdays <-  newdata %>% transform(daytype = factor(chron::is.weekend(date),levels=c(TRUE,FALSE), 
                   labels=c("Weekend","weekday")))

head(wkdays,30)

```

## Panel Plot

For whatever reason the _type = "l"_ attribute produces odd results and was not includes
```{r echo=TRUE}
library(lattice)

wkdays <-  newdata %>% transform(daytype = factor(chron::is.weekend(date),levels=c(TRUE,FALSE), 
                   labels=c("Weekend","weekday"))) %>%
                   arrange(interval,date)  %>%
                   ddply( c("date","interval","daytype"), summarise, arvsteps = mean(steps))

wkdays$arvsteps <- as.integer(wkdays$arvsteps);
head(wkdays,30)

#Create plot
xyplot(arvsteps ~ interval | daytype, data=wkdays, grid = TRUE,  layout = c(1,2),  xlab = "Intervals", ylab = "No of steps"  )


```

With _type = "l"_ included!

```{r echo=TRUE}
library(lattice)

wkdays <-  newdata %>% transform(daytype = factor(chron::is.weekend(date),levels=c(TRUE,FALSE), 
                   labels=c("Weekend","weekday"))) %>%
                   arrange(interval,date)  %>%
                   ddply( c("date","interval","daytype"), summarise, arvsteps = mean(steps))

wkdays$arvsteps <- as.integer(wkdays$arvsteps);

#Create plot
xyplot(arvsteps ~ interval | daytype, data=wkdays, grid = TRUE, type="l",  layout = c(1,2),  xlab = "Intervals", ylab = "No of steps"  )


```

